---
title: "Data Manipulation"
author: "Janish Parikh"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clear the global environment 
```{r, echo = F}
rm(list = ls())
```

# Import Required Libraries
```{r}
library(lubridate) 
```
- Lubridate provides simple functions to get and set components of a date-time, such as year(), month(), mday(), hour(), minute() and second()
```{r}
library(RColorBrewer)
```
- Provides color palettes

# Import Dataset
```{r}
citibike.May21.sample <- read.csv("~/Downloads/Datasets/citibike.May21.sample.csv")
```

Let's look at how to rename a dataframe to make it easy for us
```{r}
city <- citibike.May21.sample 
rm(citibike.May21.sample)
```

Attaching the data frame so that we no longer need to refer every attribute by typing city$
```{r}
attach(city) 
```

Let's look at the names of the attributes/colnames
```{r}
names(city) 
```

Let's look at the shape of the data at hand
```{r}
dim(city) #no of columns and rows
summary(city)
head(city,10)
str(city)
```

Let's take a deeper dive
```{r}
unique(rideable_type)
```
Since rideable_type is offering no information, we can say it's redundant in our dataset and can be removed

Removing the unwanted columns from our dataset
```{r}
city<-city[,-3] 
head(city,1)
```

```{r}
class(started_at)
```
It's character so let's change that to timestamp type so that we can find the duration of the ride
```{r}
city$started_at<-as.POSIXct(started_at)
city$ended_at<-as.POSIXct(ended_at)
class(city$started_at)
unclass(city$started_at) #See how it's actually stored inside, classes can be thought of as containers
```
```{r}
detach(city)
```
```{r}
city$duration <- as.integer(city$ended_at-city$started_at)
head(city[,c(3,4,14)])
```
```{r}
table(city$start_station_name)
```
```{r}
table(city$end_station_name)
dim(unique(city$start_station_name))-dim(unique(city$end_station_name))
```
```{r}
table(cut(city$duration, breaks = 5))
table(cut(city$duration, breaks = c(0,300,600,900,1800)))
table(cut(city$duration, breaks = c(0,300,600,1800,max(city$duration, na.rm = T)), labels =c("Small","Medium", "Long","Very Long") ))
```
Let's add this levels category to our data frame
```{r}
city$duration_type<-cut(city$duration, breaks = c(0,300,600,1800,max(city$duration, na.rm = T)), labels =c("Small","Medium", "Long","Very Long"))
cat("\014")
head(city[,c(3,4,14,15)],1)
```

Let's plot the data
```{r}
barplot(table(city$duration_type), 
        col = brewer.pal(4,"Dark2"), 
        las =1, 
        main="Barplot of Ride Durations", 
        xlab = "Ride Type based on Duration", 
        ylab = "Frequency of rides", 
        legend.text= TRUE,
        args.legend = c("Small:0-5 min", "Medium:5-10 min","Long:10-30 min", "Very Long>30 min"))
```
Let's explore the lubridate package further
- Combining two columns
- The y~x is used to say you plot y as a function of x
- Divide the duration by 60 to convert sec to min
- hour() converts date-time values in 'started_at' to hour 
```{r}
boxplot(city$duration/60 ~ hour(city$started_at), 
        xlab = "Hour ride started", 
        ylab = "Ride duration",
        main = "Boxplot of Hour ride started v/s the ride duration",
        ylim=c(0,100), 
        las =1,
        col = c(brewer.pal(8,"Dark2")))
```

Let's explore subset
```{r}
city_subset_time <- subset(city, select = c("started_at","ended_at","duration","duration_type"))
head(city_subset_time)
head(city[,c(3,4,14,15)]) #You can see that the same can be achieved by just slicing the indexes
```

```{r}
city_subset_casual <- subset(city,member_casual=="casual")
head(city_subset_casual)
```

The same results can also be obtained using this
```{r}
city_subset_casual1<- city[which(city$member_casual=="casual"),]
head(city_subset_casual1)
```

As you can see both the results are the same

Let's explore tapply, sapply, lapply, apply
find average ride duration for each hour
```{r}
city_houravg<-tapply(city$duration,hour(city$started_at),mean)
View(city_houravg)
```

Now let's see how data manipulation actually hepled us to gain actionable intelligence
# (1) Ride Length by Day for Casual Riders
```{r}
boxplot((city_subset_casual$duration/60)~day(city_subset_casual$started_at),
        las=2, 
        xlab = "Day", 
        ylab = "Duration (mins)", 
        main="Ride Length by Day for Casual Riders",
        ylim=c(0,100),
        col = c(brewer.pal(11,"Spectral")))
```

# (2) Average Ride Length for each Hour
```{r}
barplot(city_houravg, main="Average Ride Length for each Hour", font.main=4, col.main="red", xlab="Hour", ylab="Average Ride length in seconds", col = brewer.pal(12,"Set3"))
```

# (3) Plot Ride Length by Rider Type
```{r}
mosaicplot(city$member_casual~city$duration_type, shade=F,xlab= "Rider Type", ylab = "Ride Type", col = brewer.pal(4,"Set1"), main="Ride Length Type per Rider Type")
```

# (4) scatter plot of latitudes and longitudes
```{r}
plot(city_subset_casual$end_lat, 
     city_subset_casual$end_lng, 
     col = c("orange","cyan"))
```