---
title: "Basics of R and R Studio"
author: "Janish Parikh"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

Shortcut to comment/uncomment - Ctrl + Shift + C

To run the below commands - Select the line to run and click on "Run" (Present on the top right corner of this window)
Note : It is better to write each command on your own for practice and learning.

# Import library
Readr - Read rectangular text data

```{r}
library(readr)
```

# Import Dataset
Note : Import your dataset using the name - "automobiles"
Click on the "Import Dataset" and select "from Text(readr)" and load your dataset.
```{r}
automobiles <- read.csv("~/Downloads/Datasets/automobiles.csv")
head(automobiles)
```
# Understanding the data at hand
Structure of the object/dataset
```{r}
str(automobiles)
```
Length of the object/dataset
```{r}
length(automobiles)
```
Class of the object/dataset - High level
```{r}
class(automobiles)
```
Type of the object/dataset - Low-level
```{r}
typeof(automobiles)
```
Name of the object/dataset
```{r}
names(automobiles)
```
Note : If you type "$" after any dataframe it will show you the different variables present in the dataframe

Try - class(automobiles$normalized.losses) and typeof(automobiles$normalized.losses), as two separate commands

See what happens
```{r}
class(automobiles$normalized.losses)
```

```{r}
typeof(automobiles$normalized.losses)
```

Do you understand why this happens?

# Datatypes

- character
- numeric (real or decimal)
- integer
- logical
- complex

```{r}
a <- "a"
b <- 2.5
c <- 6
d <- TRUE
e <- 1+4i
```
You will see them appear on the Workspace as and when you store values in variables
The Workspace is present on the right side

How to see what is present in each variable? - Just type the variable name in the console
For example : Type b in the console and press enter. It will display 2.5
Note : This can be done for any of the other variables. Try doing that for other variables and see what is displayed.


# Data structures

Note : "c()" is a combine function. It combines the information inside it and stores it in the variable.

## Vector
```{r}
f <- c(1,2,5.3,6,-2,4) # numeric vector
g <- c("one","two","three") # character vector
h <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
```

## Matrix
```{r}
i<-matrix(1:20, nrow=5,ncol=4)
```

## Dataframes 
Main idea is that each value can be of differnt modes(numeric, character, factor ,.etc)
```{r}
j <- c(1,2,3,4)
k <- c("red", "white", "red", NA)
l <- c(TRUE,TRUE,TRUE,FALSE)
```

How to create a dataframe using different modes
```{r}
mydata <- data.frame(j,k,l)
```

Giving column names for each of the three variables. Think of naming each column with a header name
```{r}
names(mydata) <- c("ID","Color","Passed") # variable names
```

## Lists
Ordered collection of objects
Note : Lists is not used a lot. Just understand what a list is.
```{r}
w <- list(name="Name", numbers = j)
```
## Factors 

Tells R that the variable is nominal. Used for categorical variables.
Note : Do not concentrate on what rep() as a function does. Concentrate on WHY gender is a factor in R.
```{r}
gender <- c(rep("male",20), rep("female", 30)) 
```

This is how a categorical variable is converted to a factor (data structure) in R
```{r}
gender <- factor(gender) 
```

Display the contents of gender
```{r}
gender
```

# Plots

## Histogram plot
Used to see the distribution (Continuous data)
```{r}
hist(automobiles$compression.ratio)
```

## Box plot
To show the distribution of data (Min, First Quartile, Median, Third Quartile, Max)
```{r}
boxplot(automobiles$compression.ratio ~ automobiles$make)
```

## Bar Plot
Used to compare variables (For categorical data)
```{r}
table(automobiles$make)
barplot(table(automobiles$make))
```

## Mosaic Plot
Represents a contingency table (Distribution of two variables against each other) 
[Box size represents the cell frequencies]
```{r}
mosaicplot(automobiles$make ~ automobiles$fuel.type)
```

# How to use help
```{r}
?barplot
??bar  
```